Making Sequential Programs Deterministic

### 2019-03-26 design.tex
Fixes - removing @ from types

The underlying checker-framework on which we build our determinism checker has support for
polymorphic annotations. When a user writes a polymorphic annotation on a method signature or a type parameter,
it indicates that it could resolve to any type in the type system depending on how it is used.
In the determinism checker, we define a polymorphic annotation \<PolyDet>.
One of the most common locations for a polymorphic annotation is at a method signature.
Consider the following method declaration annotated with polymorphic annotations:
\begin{verbatim}
@PolyDet int foo(@PolyDet String param1, @PolyDet boolean param2) { }
\end{verbatim}
There are two valid instantiations for this declaration:
\begin{enumerate}
    \item @NonDet int foo(@NonDet String param1, @NonDet boolean param2) { }
    \item @Det int foo(@Det String param1, @Det boolean param2) { }
\end{enumerate}
This indicates that method \<foo> can be called with arguments having any of the \codeid{@Det or @NonDet} type annotations.
%   (\<@OrderNonDet> is not allowed because it is invalid on primitive types).
 \<@PolyDet> resolves to the least upper bound of
the actual types on arguments. For instance, if this method is called as \codeid{foo(@Det String arg1, @NonDet boolean arg2)}, the 
checker framework resolves the method declaration as \codeid{@NonDet int foo(@NonDet String param1, @NonDet boolean param2)}
causing the return type to have the type annotation \<@NonDet>.

@@

The possibility of mutation allows us to add to the \<@Det List> at a
\<@NonDet> index, which is unsound.

@@

Some examples of valid types are:
\begin{itemize}
    \item \codeid{@NonDet\ \ \ \ \ \ List<@NonDet\ \ \ \ \ \ Integer>}
    \item \codeid{@Det\ \ \ \ \ \ \ \ \ List<@Det\ \ \ \ \ \ \ \ \ Integer>}
    \item \codeid{@OrderNonDet List <@OrderNonDet Set<...>\relax >}
    \item \codeid{@OrderNonDet List <@Det String>}
\end{itemize}

These types are invalid:
\begin{itemize}
    \item \codeid{@OrderNonDet Set <@NonDet\ \ \ \ \ \ Integer>}
    \item \codeid{@NonDet\ \ \ \ \ \ List<@Det\ \ \ \ \ \ \ \ \ Integer>}
    \item \codeid{@Det\ \ \ \ \ \ \ \ \ List<@OrderNonDet Set<...>\relax >}
\end{itemize}

Similarly, the (determinism) type of an array must be a supertype or equal to
the type of the component type (see Figure~\ref{fig-determinism-collections}).
As with collections, \codeid{@NonDet} arrays of \codeid{@Det} or \codeid{@OrderNonDet}
elements are not allowed (Rules $\sc{\mathit{ORDERNONDET \ ARRAY}}$ and $\sc{\mathit{NONDET \ ARRAY}}$ in Figure~\ref{type-validity}).

@@

A collection (\<List, Set, or a Map>) annotated as \<@OrderNonDet> has the following properties:
\begin{enumerate}
    \item The individual elements extracted by iterating over the collection get the type \<NonDet>.
    \item Accessing properties sich as \<size()> or querying if the collection \<isEmpty()> will return types
    that are annotated as \<Det> as they do not depend on the iteration order. 
\end{enumerate}

A user defined type may be annotated as \<OrderNonDet> if and only if it
is a subtype of Collection or Iterator.
For example, if a user defines a type as
\begin{verbatim}
public class TestUserCollection<E> extends ArrayList<E> {...}
\end{verbatim}
Writing \codeid{@OrderNonDet TestUserCollection<@Det Integer>} is valid.\\
Writing \codeid{@OrderNonDet TestUserCollection<@NonDet Integer>} is invalid
because the type of the type parameter (\codeid{@NonDet}) is not a subtype
of the user defined Collection type (\codeid{@OrderNonDet}).

User defined types that are not a subtype of Collection or Iterator
may not be annotated as \codeid{@OrderNonDet}.

@@

Recall that the determinism checker does not permit \<@OrderNonDet> on a type that isn't a collection, iterator, or an array.
As a consequence, resolution of polymorphic type annotations at method signatures could have undesirable effects.
Suppose a method declaration is annotated as \codeid{@PolyDet int size(@PolyDet List<T> list)}. Invoking this method
as \codeid{size(@OrderNonDet List<@Det String>)} will result in \<@PolyDet> getting resolved to \<@OrderNonDet> which is invalid
on the return type \<int>. We circumvent this scenario by introducing two variants to the \<@PolyDet> annotation, \<@PolyDet("up")>
and \<@PolyDet("down")>.

@@

The semantics of these variants are as follows:
\begin{itemize}
    \item \<@PolyDet("up")>: Same as \<@PolyDet> except when it resolves to \<@OrderNonDet>, it is replaced with \<@NonDet>.
    \item \<@PolyDet("down")>: Same as \<@PolyDet> except when it resolves to \<@OrderNonDet>, it is replaced with \<@Det>.
\end{itemize} 
The modifiers "up" and "down" are indicative of the subtyping relationship of \<@OrderNonDet> with \<@NonDet> and \<@Det> respectively
in the determinism type hierarchy.
Annotating \<size()> as \codeid{@PolyDet("down") int size(@PolyDet List<T> list)} will only allow the following type resolutions:
\begin{itemize}
    \item \codeid{@Det int size(@Det List<T> list)}
    \item \codeid{@Det int size(@OrderNonDet List<T> list)}
    \item \codeid{@NonDet int size(@NonDet List<T> list)}
\end{itemize}

\begin{figure}
    $\infer[\sc{\mathit{POLY RESOLUTION}}]{\pi_r \ \beta_r method(\overline{\pi_i \ \beta_i})\{\ \}}{\forall K \in \{\mathit{NonDet, OrderNonDet, Det\}} : K \ \beta_r method(\overline{K \ \beta_i})\{\ \} & \pi_r, \pi_i \in \{\mathit{PolyDet}, \mathit{PolyDet}\uparrow, \mathit{PolyDet}\downarrow    \}}$
\end{figure}

@@

In order to provide soundness guarantees, we must ensure that a collection that is
declared as \<@Det> is not side-effected in non-deterministic ways. Consider the following annotation 
on the List add method: \codeid{public void add(@PolyDet ArrayList<E> this, @PolyDet int index, E element)}.
Suppose a client calls this method as follows:
\begin{verbatim}
@Det ArrayList<@Det String> list;
@NonDet int random;
@Det String str;
list.add(random, str);
\end{verbatim}
This will result in the instantion \codeid{public void add(@NonDet ArrayList<E> this, @NonDet int index, E element)} and
the method invocation will type check. This is problematic because even though \<list> was declared to be \<@Det>,
the type checker allows the addition of an element at a \<@NonDet> location violating our determinism guarantees.
We prevent this behavior by introducing another valiant of \<@PolyDet>, \<@PolyDet("use")> that does not affect
type instantiations. This is especially useful in preventing methods from non-deterministically modifying the state
of a deterministic receiver. To have this effect, the library method \<ArrayList.add> is annotated as:
\codeid{public void add(@PolyDet ArrayList<E> this, @PolyDet("use") int index, E element)}.
Since \<@PolyDet("use")> has no effect on type instantiations, calling this method on a \<@Det List>
will result in the method declaration being resolved as \codeid{public void add(@Det ArrayList<E> this, @Det int index, E element)}
irrespective of the type of \<index>, thereby preventing undesired side-effects.

\subsection{Discussion on defaulting rules}\label{defaulting}
Every type system in the checker framework must define a default type annotation. This annotation will be applied
to every location that isn't explicitly annotated. In our determinism checker, we chose \<@PolyDet> as the default for all 
method parameters and return types including constructors.
The rationale for this design choice was to not make our type checker too restrictive by making everything \<@Det>.  This implies that
the checker still allows some non-determinism as long as it doesn't break any determinism specifications explicitly written by the programmer.

@@

\subsection{Dataflow analysis and type refinement}\label{dataflow}
The checker framework performs automatic type inference for local types within method bodies.
It also performs dataflow analysis and refines the types of these locals, thereby reducing the annotation burden 
on the programmer. Consider the example below:
\begin{verbatim}
void foo(@Det int x) {
    @NonDet int y = x;    // Type of y gets refined to @Det.
    @Det int z = y;       //  No error here.
}
\end{verbatim}
In the example above, even though the local variable \codeid{y} was declared to be \codeid{@NonDet}, it gets
type refined to \codeid{@Det} after the first assignment statement. As the result, the assignment of \codeid{y}
to a \codeid{@Det} variable \codeid{z} is sound. In the following example,
\begin{verbatim}
void foo(@NonDet Object x, @Det Object y) {
    x = y;                // No type refinement for x.
    @Det Object z = x;    // Error.
}
\end{verbatim}
the checker reports a warning at the second assignment statement of \codeid{foo}. Type refinement only applies to locals. Refining non-locals can result in side effects making the analysis unsound.

With the above rules, the checker framework will automatically annotate a local collection with type \<@NonDet> and its
type arguments will get the type qualifier \<@Det>. For example, a locally declared \codeid{List} of \codeid{Strings} will be automatically annotated
as \codeid{@NonDet List<@Det String>}. Recall from~\ref{collection-rules} that the determinism checker does not allow this type on
any collection. While this is still sound, it will result in the determinism checker reporting a lot of errors because every locally declared 
collection gets annotated with this invalid type. To avoid this situation, we make an exception to the defaulting rules for local collections
and annotate their type parameters also with \<@NonDet>. So a local list of Strings gets the default annotation of \codeid{@NonDet List<@NonDet String>}. While this defaulting eliminates the possibility of automatically annotating locals collections with an invalid
type, it could result in a high number of false positives. In the following code snippet,
\begin{verbatim}
void test(@Det List<@Det String> argList) {
    List<String> localList = argList;
}    
\end{verbatim}
the checker would report an error at the assignment statement since the inferred type for \codeid{localList} is 
\codeid{@NonDet List<@NonDet> String} and since collections types are invariant.
We therefore recommend that programmers explicitly annotate
all local variables that are collections to reduce false positives. The same reasoning and recommendation applies to arrays.

@@

    \caption{Sorting type refinement rules}
    \label{fig-sorting}
\end{figure}
The checker type refines a receiver of type \codeid{@OrderNonDet List} to a \codeid{@Det List} if all the
type argument of this \codeid{List} have the type annotation \codeid{@Det} (Rule LIST SORT in Figure~\ref{fig-sorting}). Similar rules apply for \codeid{Arrays.sort()} and \codeid{Collections.sort()}.
As per these rules, if \codeid{sort()} is called on a receiver of type \codeid{@OrderNonDet List<@Det Set<@Det String>>}, it will
be type refined to \codeid{@Det List<@Det Set<@Det String>>}. However a receiver of type \codeid{@OrderNonDet List<@OrderNonDet Set<@Det String>>} will not be type refined.

>>>>

The underlying checker-framework on which we build our determinism checker has support for
polymorphic annotations. When a user writes a polymorphic annotation on a method signature or a type parameter,
it indicates that it could resolve to any type in the type system depending on how it is used.
In the determinism checker, we define a polymorphic annotation \<PolyDet>.
One of the most common locations for a polymorphic annotation is at a method signature.
Consider the following method declaration annotated with polymorphic annotations:
\begin{verbatim}
PolyDet int foo(PolyDet String param1, PolyDet boolean param2) { }
\end{verbatim}
There are two valid instantiations for this declaration:
\begin{enumerate}
    \item NonDet int foo(NonDet String param1, NonDet boolean param2) { }
    \item Det int foo(Det String param1, Det boolean param2) { }
\end{enumerate}
This indicates that method \<foo> can be called with arguments having any of the \codeid{Det or NonDet} type annotations.
%   (\<@OrderNonDet> is not allowed because it is invalid on primitive types).
 \<PolyDet> resolves to the least upper bound of
the actual types on arguments. For instance, if this method is called as \codeid{foo(Det String arg1, NonDet boolean arg2)}, the 
checker framework resolves the method declaration as \codeid{NonDet int foo(NonDet String param1, NonDet boolean param2)}
causing the return type to have the type annotation \<NonDet>.

@@

The possibility of mutation allows us to add to the \<Det List> at a
\<NonDet> index, which is unsound.

@@

Some examples of valid types are:
\begin{itemize}
    \item \codeid{NonDet\ \ \ \ \ \ List<NonDet\ \ \ \ \ \ Integer>}
    \item \codeid{Det\ \ \ \ \ \ \ \ \ List<Det\ \ \ \ \ \ \ \ \ Integer>}
    \item \codeid{OrderNonDet List <OrderNonDet Set<...>\relax >}
    \item \codeid{OrderNonDet List <Det String>}
\end{itemize}

These types are invalid:
\begin{itemize}
    \item \codeid{OrderNonDet Set <NonDet\ \ \ \ \ \ Integer>}
    \item \codeid{NonDet\ \ \ \ \ \ List<Det\ \ \ \ \ \ \ \ \ Integer>}
    \item \codeid{Det\ \ \ \ \ \ \ \ \ List<OrderNonDet Set<...>\relax >}
\end{itemize}

Similarly, the (determinism) type of an array must be a supertype or equal to
the type of the component type (see Figure~\ref{fig-determinism-collections}).
As with collections, \codeid{NonDet} arrays of \codeid{Det} or \codeid{OrderNonDet}
elements are not allowed (Rules $\sc{\mathit{ORDERNONDET \ ARRAY}}$ and $\sc{\mathit{NONDET \ ARRAY}}$ in Figure~\ref{type-validity}).

@@

A collection (\<List, Set, or a Map>) annotated as \<OrderNonDet> has the following properties:
\begin{enumerate}
    \item The individual elements extracted by iterating over the collection get the type \<NonDet>.
    \item Accessing properties sich as \<size()> or querying if the collection \<isEmpty()> will return types
    that are annotated as \<Det> as they do not depend on the iteration order. 
\end{enumerate}

A user defined type may be annotated as \<OrderNonDet> if and only if it
is a subtype of Collection or Iterator.
For example, if a user defines a type as
\begin{verbatim}
public class TestUserCollection<E> extends ArrayList<E> {...}
\end{verbatim}
Writing \codeid{OrderNonDet TestUserCollection<Det Integer>} is valid.\\
Writing \codeid{OrderNonDet TestUserCollection<NonDet Integer>} is invalid
because the type of the type parameter (\codeid{NonDet}) is not a subtype
of the user defined Collection type (\codeid{OrderNonDet}).

User defined types that are not a subtype of Collection or Iterator
may not be annotated as \codeid{OrderNonDet}.

@@

Recall that the determinism checker does not permit \<OrderNonDet> on a type that isn't a \<Collection>, \<Iterator>, or an array.
As a consequence, resolution of polymorphic type annotations at method signatures could have undesirable effects.
Suppose a method declaration is annotated as \codeid{PolyDet int size(PolyDet List<T> list)}. Invoking this method
with an argument of type \codeid{OrderNonDet List<Det String>} will result in \<PolyDet> getting resolved to \<OrderNonDet> which is invalid
on the return type \<int>. We circumvent this scenario by introducing two variants to the \<PolyDet> annotation, \<PolyDet$\uparrow$>
and \<PolyDet$\downarrow$>.

@@

The semantics of these variants are as follows:
\begin{itemize}
    \item \<PolyDet$\uparrow$>: Same as \<PolyDet> except when it resolves to \<OrderNonDet>, it is replaced with \<NonDet>.
    \item \<PolyDet$\downarrow$>: Same as \<PolyDet> except when it resolves to \<OrderNonDet>, it is replaced with \<Det>.
\end{itemize} 
The modifiers $\uparrow$ and $\downarrow$ are indicative of the subtyping relationship of \<OrderNonDet> with \<NonDet> and \<Det> respectively
in the determinism type hierarchy.

Figure~\ref{fig-poly-resolution} presents the rule for type checking of methods annotated with \<PolyDet> or its variants.

Annotating \<size()> as \codeid{PolyDet$\downarrow$ int size(PolyDet List<T> list)} will only allow the following type resolutions:
\begin{itemize}
    \item \codeid{Det int size(Det List<T> list)}
    \item \codeid{Det int size(OrderNonDet List<T> list)}
    \item \codeid{NonDet int size(NonDet List<T> list)}
\end{itemize}

\begin{figure}
    $\infer[\sc{\mathit{POLY RESOLUTION}}]{\pi_r \ \beta_r method(\overline{\pi_i \ \beta_i})\{\ \}}{\forall K \in \{\mathit{NonDet, OrderNonDet, Det\}} : K \ \beta_r method(\overline{K \ \beta_i})\{\ \} & \pi_r, \pi_i \in \{\mathit{PolyDet}, \mathit{PolyDet}\uparrow, \mathit{PolyDet}\downarrow    \}}$
    \caption{Polymorphic resolution rule}
    \label{fig-poly-resolution}
\end{figure}

@@

In order to provide soundness guarantees, we must ensure that a collection that is
declared as \<Det> is not side-effected in non-deterministic ways. Consider the following annotation 
on the List add method: \codeid{public void add(PolyDet ArrayList<E> this, PolyDet int index, E element)}.
Suppose a client calls this method as follows:
\begin{verbatim}
@Det ArrayList<@Det String> list;
@NonDet int random;
@Det String str;
list.add(random, str);
\end{verbatim}
This will result in the instantion \codeid{public void add(NonDet ArrayList<E> this, NonDet int index, E element)} and
the method invocation will type check. This is problematic because even though \<list> was declared to be \<Det>,
the type checker allows the addition of an element at a \<NonDet> location violating our determinism guarantees.
We prevent this behavior by introducing another valiant of \<PolyDet>, \<PolyDet("use")> that does not affect
type instantiations. This is especially useful in preventing methods from non-deterministically modifying the state
of a deterministic receiver. To have this effect, the library method \<ArrayList.add> is annotated as:
\codeid{public void add(PolyDet ArrayList<E> this, PolyDet("use") int index, E element)}.
Since \<PolyDet("use")> has no effect on type instantiations, calling this method on a \<Det List>
will result in the method declaration being resolved as \codeid{public void add(Det ArrayList<E> this, Det int index, E element)}
irrespective of the type of \<index>, thereby preventing undesired side-effects.

\subsection{Discussion on defaulting rules}\label{defaulting}
Every type system in the checker framework must define a default type annotation. This annotation will be applied
to every location that isn't explicitly annotated. In our determinism checker, we chose \<PolyDet> as the default for all 
method parameters and return types including constructors.
The rationale for this design choice was to not make our type checker too restrictive by making everything \<Det>.  This implies that
the checker still allows some non-determinism as long as it doesn't break any determinism specifications explicitly written by the programmer.

@@

\subsection{Dataflow analysis and type refinement}\label{dataflow}
The checker framework performs automatic type inference for local types within method bodies.
It also performs dataflow analysis and refines the types of these locals, thereby reducing the annotation burden 
on the programmer. Consider the example below:
\begin{verbatim}
void foo(@Det int x) {
    @NonDet int y = x;    // Type of y gets refined to @Det.
    @Det int z = y;       //  No error here.
}
\end{verbatim}
In the example above, even though the local variable \codeid{y} was declared to be \codeid{NonDet}, it gets
type refined to \codeid{Det} after the first assignment statement. As the result, the assignment of \codeid{y}
to a \codeid{Det} variable \codeid{z} is sound. In the following example,
\begin{verbatim}
void foo(@NonDet Object x, @Det Object y) {
    x = y;                // No type refinement for x.
    @Det Object z = x;    // Error.
}
\end{verbatim}
the checker reports a warning at the second assignment statement of \codeid{foo}. Type refinement only applies to locals. Refining non-locals can result in side effects making the analysis unsound.

With the above rules, the checker framework will automatically annotate a local collection with type \<NonDet> and its
type arguments will get the type qualifier \<Det>. For example, a locally declared \codeid{List} of \codeid{Strings} will be automatically annotated
as \codeid{NonDet List<Det String>}. Recall from~\ref{collection-rules} that the determinism checker does not allow this type on
any collection. While this is still sound, it will result in the determinism checker reporting a lot of errors because every locally declared 
collection gets annotated with this invalid type. To avoid this situation, we make an exception to the defaulting rules for local collections
and annotate their type parameters also with \<NonDet>. So a local list of Strings gets the default annotation of \codeid{NonDet List<NonDet String>}. While this defaulting eliminates the possibility of automatically annotating locals collections with an invalid
type, it could result in a high number of false positives. In the following code snippet,
\begin{verbatim}
void test(@Det List<@Det String> argList) {
    List<String> localList = argList;
}    
\end{verbatim}
the checker would report an error at the assignment statement since the inferred type for \codeid{localList} is 
\codeid{NonDet List<NonDet> String} and since collections types are invariant.
We therefore recommend that programmers explicitly annotate
all local variables that are collections to reduce false positives. The same reasoning and recommendation applies to arrays.

@@

    \caption{Sorting type refinement rules}
    \label{fig-sorting}
\end{figure}
The checker type refines a receiver of type \codeid{OrderNonDet List} to a \codeid{Det List} if all the
type argument of this \codeid{List} have the type annotation \codeid{Det} (Rule LIST SORT in Figure~\ref{fig-sorting}). Similar rules apply for \codeid{Arrays.sort()} and \codeid{Collections.sort()}.
As per these rules, if \codeid{sort()} is called on a receiver of type \codeid{OrderNonDet List<Det Set<Det String>>}, it will
be type refined to \codeid{Det List<Det Set<Det String>>}. However a receiver of type \codeid{OrderNonDet List<OrderNonDet Set<Det String>>} will not be type refined.
